---
title: "Bovet & Raymond Challenge"
format: html
editor: visual
---

```{r}
rm(list=ls())
library(tidyverse)
library(dplyr)
library(broom)
library(cowplot)
```

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/BovetAndRaymondData.csv"
d <- read_csv(f, col_names = TRUE)
names(d)
View(d)

ggplot(data = d, aes(x = Century, y = `Corrected WHR`)) +
  geom_point() +
  geom_smooth(method = "lm")

m <- lm(`Corrected WHR` ~ Century, data = d)
summary(m)
#Residuals:
#      Min        1Q    Median        3Q       Max 
#-0.074101 -0.021963 -0.001957  0.021954  0.069009 

#Coefficients:
#              Estimate Std. Error t value Pr(>|t|)    
#(Intercept)  0.7483684  0.0049591 150.909  < 2e-16 ***
#Century     -0.0009634  0.0002990  -3.222  0.00147 

observed_coef <- tidy(m) |> 
  filter(term == "Century") |>
  pull(estimate)
observed_coef #-0.000963389
```

Permutation Test - decouple relationship between century and WHR and conduct the regression model again

```{r}
n_perm <- 1000
perm_coef <- vector() #in this case, coef of interest is the slope

for(i in 1:n_perm){
  d_perm <-  d #make a fresh copy for each iteration
  d_perm |> mutate(Century = sample(Century)) |> #shuffle relationships
    perm_coef[[i]] <- lm(`Corrected WHR` ~ Century, data = d_perm) |>
      tidy() |>
      filter(term == "Century") |> #take the coefficient out of the return table
      pull(estimate)
}
#NOT WORKING
```

Calculate a two-tailed P value How often does a permuted coefficient exceed our observed coefficient in magnitude?

```{r}
p_val <- mean(abs(perm_coef) >= abs(observed_coef))
cat("Permutation p-value:", p_val, "\n")
#Permutation p-value: 9e-04 DONT TRUST

#Compare to classic parametric p-value that was returned in the original model
observed_coef_p <- tidy(m) |> 
  filter(term == "Century") |>
  pull(p.value)
cat("Parametric p-value:", observed_coef_p, "\n")
#Parametric p-value: 0.001470502 
```

Bootstrapping Confidence Interval The above process (permutation) gives us a null coefficient that we can compare to our observed coefficient. Bootstrapping creates a real coefficient from our data by taking samples from the data over and over again. Sampling occurs with replacement

```{r}
nboot <- 1000
boot_coef <- vector()
for(i in 1:nboot) {
  d_boot <- d[sample(nrow(d), replace = TRUE), ] #resample rows w replacement
  boot_coef[[i]] <- lm(`Corrected WHR` ~ Century, data = d_boot) |>
    tidy() |> 
    filter(term == "Century") |>
    pull(estimate)
}

hist(boot_coef)

ci_boot <- quantile(boot_coef, c(0.025, 0.975))
cat("Bootstrap 95% CI:", ci_boot[[1]], "to", ci_boot[[2]], "\n")

##cat("Parametric 95% CI:", ) #check back later to finish

tidy(m, conf.int = TRUE) #will return CI in model summary for parametric
```

How to check your model for interference

-   Linearity of relationship between variables

-   Independence of residuals

-   Normality of residuals

-   Equality of variance (AKA homoscedasticity")

L, N, and E cam be evaluated using residual analysis

## Checking Residuals - Kamilar & Cooper Data

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/KamilarAndCooperData.csv"
dd <- read_csv(f, col_names = TRUE)

#Run linear models of the relationship using different log adjustments
p1 <- ggplot(data = dd, aes(x = Body_mass_female_mean, y = MaxLongevity_m)) +
  geom_point(na.rm = TRUE)
p2 <- ggplot(data = dd, aes(x = log(Body_mass_female_mean), y = MaxLongevity_m)) +
  geom_point(na.rm = TRUE)
p3 <- ggplot(data = dd, aes(x = log(Body_mass_female_mean), y = log(MaxLongevity_m))) +
  geom_point(na.rm = TRUE)

#using cowplot
plot_grid(p1, p2, p3, nrow = 1)
```

Create linear models of the above relationships, extract residuals, and make histograms. Are they normal? 

```{r}
m1 <- lm(MaxLongevity_m ~ Body_mass_female_mean, data = dd)
m2 <- lm(MaxLongevity_m ~ log(Body_mass_female_mean), data = dd)
m3 <- lm(log(MaxLongevity_m) ~ log(Body_mass_female_mean), data = dd)

pm1 <- ggplot(data = NULL, aes(x = m1$model$Body_mass_female_mean, y = m1$residuals)) +
  geom_point(na.rm = TRUE)
#should be centered around 0 because we have fit them around a linear relationship
#looking to see the relationship between x value and residuals, we expect to see a shotgun spatter if there is no relationship
pm2 <- ggplot(data = NULL, aes(x = m2$model$`log(Body_mass_female_mean)`, y = m2$residuals)) +
  geom_point(na.rm = TRUE)
#closer to appearance of randomness
pm3 <- ggplot(data = NULL, aes(x = m3$model$`log(Body_mass_female_mean)`, y = m3$residuals)) +
  geom_point(na.rm = TRUE)
#also appears random, better centering aorund 0 on y axis

rm1 <- histogram(m1$residuals) #looks semi-normal
rm2 <- histogram(m2$residuals) #normalish, positive tail
rm3 <- histogram(m3$residuals) #normallest 
plot_grid(pm1, rm1, pm2, rm2, pm3, rm3, nrow = 3)

```

Checking normality using QQ Plots
```{r}
#looking for the most points to be inside the 95% CI, closer to linear 
car::qqPlot(m1$residuals)
car::qqPlot(m2$residuals)
car::qqPlot(m3$residuals) #no points outside of CI, best model for describing this data? 

#can also us ggpubr::ggqqplot()
```

Using statistical tests to determine normality of residuals -> Shapiro-Wilkes Test
```{r}
#could our vector of values come from a normal distribution? 
#when pval is < 0.05, reject that our residuals are normal (H0)

shapiro.test(m1$residuals) #W = 0.96352, p-value = 0.0008238
shapiro.test(m2$residuals) #W = 0.97515, p-value = 0.01131
shapiro.test(m3$residuals) #W = 0.99319, p-value = 0.7403
#M3 has the most normal distribution, in fact, it is above significance so we cannot reject the H0 that the vector comes from a normal distribution

```