---
title: "Part 2"
format: html
editor: visual
---

# Part 2: Statistics and Inference

Packages

```{r}

install.packages("mosaic")
install.packages("radiant")
install.packages("moments")

library(mosaic)
library(radiant)
library(moments)
library(dplyr)
library(tidyverse)

```

# Class 10: Intro to Statistics

## Statistics - Key Ideas

*Population*: all the individuals of interest

*Sample*: a representative subset of the population that allows us to make inferences about the population

The goal of statistics is to make a robust claim about a population based on a sample of data points. This is biased in some ways, because all samples are \< populations. We can use bias-adjustment methods on the sample to correct for this, or other, potential biases.

*Parameter*: descriptions of the variable of interest in the larger population based on statistical inference, estimates depend on robusticity of statistics

*Statistic*: some function of the data alone, a combination based on a finite amount of data, can only describe the sample - not the population

What kinds of statistics are we interested in?

*Central tendency*: What is the mean/median/central value of our data?

-   mean (average)
-   median (central numerical point)
-   mode (most common value)
-   harmonic mean (average of 1/each observation, inverse mean, less susceptible to outliers)

*Spread*: How much do the non-central values deviate from the central tendency?

-   mean deviation (avg. of abs. values of deviations from median)

-   mean squared deviation (avg. of squared deviations from mean) AKA variance

-   standard deviation (square root of variance)

-   inter-quartile range (value range between 25% and 75% of observations)

-   range (maximum - minimum value)

*Population vs. Sample spread*

-   population variance = a parameter, where N = population size

$$\frac{SS}{N} = \sigma^2$$

-   population sd = $\sigma$

-   sample variance = a statistic = SS / n-1 = s\^2, where n = sample size

$$\frac{SS}{n-1} = s^2$$

-   sample sd = s

*Shape*: the evenness or skewness of variation within the distribution

-   skewness: characterizes the assymetry of a distribution $$
    g_1 = (1/n) \sum_i^n (\frac{x_i - \bar{x}}{s})^3
    $$
-   kurtosis: characterizes the peakedness or flatness of a distribution (compared with noramlity) $$
    g_2 = (1/n) \sum_i^n (\frac{x_i - \bar{x}}{s})^4
    $$ \## Example Code for Spread

```{r}
#population variance
sum(((v-mean(v))^2)/length(v))

#sample variance
sum(((v-mean(v))^2)/(length(v)-1))
var(v)

#population standard deviation
radiant::sdpop #check this one

#sample standard deviation
sd(v)

#skewness
sum(((v-mean(v))/sd(v))^3)/(length(v)-1)
moments::skewness(v)
```

## Challenge: Calculating Statistics Write your own functions for the following:

-   calculate the variance for a vector of values representing an entire population of measurements
-   calculate the variance for a vector representing a sample \_ calculate skewness

```{r}

pop_var <- function(v){
  sum(((v-mean(v))^2)/length(v))
}

#not working vvv
sam_var <- function(v){
  sum(((v-mean(v))^2)/(length(v)-1))
}

v <- c(1,2,3,4,5)
pop_var(v) #2
sam_var(v) #2.5

skew_var <- function(v){
  sum(((v-mean(v))/sd(v))^3)/(length(v)-1)
}

skew_var(v) #0
v2 <- c(1, 1, 2, 3, 4, 5)
skew_var(v2) #0.25
```

## Class 11: Distributions

Some statistics are estimates of parameters of well-defined distributions in classical statistical inference.

Parameters such as $\mu$ and $\sigma$ of the Gaussian (normal) distribution, or $\lambda$ of the Poisson distribution define these distribution functions but are not statistics.

Normal Distributions

```{r}
#plot a normal distribution
library(mosaic)
mu <- 10 #mean parameter
sigma <- 2 #sd parameter
plotDist("norm", mean = mu, sd = sigma, xlab = "x", ylab = "Frequency")
#first argument defines the distribution

#random sampling from the population expressed by the normal dist
s1 <- rnorm(n=10, mean = 10, sd = 2) #pull from a normal pop with these rules, pull 10 individuals for the sample
histogram(s1)
mean(s1) #9.915
sd(s1) #1.769

s2 <- rnorm(n = 10000, mean = 10, sd = 2) #10000 individuals
histogram(s2) #looks more normal
mean(s2) #10.014
sd(s2) #2.037
```

What is the skewness of a normal distribution? What about our samples?

```{r}
#normal distributions have a skewness of 0, kurtosis of 3

s <- rnorm(n=10000, mean = 10, sd = 2)
moments::skewness(s) #0.0455, expect to be closer to 0 as dist gets more normal
moments::kurtosis(s) #3.114
```

F Distributions

```{r}
#parameters are degrees of freedom, diff than plotting a norm
df1 <- 5 #moderate right skew
df2 <- 20

plotDist("f", df1 = df1, df2 = df2, xlab = "x", ylab = "frequency")

s <- rf(n=10000, df1 = 5, df2 = 20)
histogram(s) #also right skewed
mean(s) #1.106
sd(s) #0.82
moments::skewness(s) #1.795
moments::kurtosis(s) #8.487, higher than 3 so it is more sharply peaked than the norm
```

Beta Distribution - bounded between 0 - 1

```{r}
shape1 = 5 #aka alpha
shape2 = 3 #aka beta

plotDist("beta", shape1 = shape1, shape2 = shape2)

s <- rbeta(n = 10000, shape1 = shape1, shape2 = shape2)
histogram(s) #slight left skew
mean(s) #0.627
sd(s) #0.160
moments::skewness(s) #-0.279
moments::kurtosis(s) #2.572
```

## Sampling Distribution of a Statistic

If we were to sample many times from a distribution/population and caclulate summary stastics, we will get slightly different values for each sample. We can then define a new distribution of those calculated statistics.

*Sampling distribution*: set of possible statistics that could have been generated if the data collection process were repeated many times, along with the probability of these possible values

Creating a sampling distribution for the mean

```{r}
reps <- 500 #set how many times to do the next thing
sdm <- do(reps) * mean(rnorm(n=1000, mean = 10, sd = 2)) #setting a data frame containing the resulting mean for each of 500 samples taken from the defined normal distribution
##sdm means sample distribution mean##
str(sdm) #500 observations, 1 variable
histogram(sdm$mean) #looks pretty normal
mean(sdm$mean) #9.979, really close to 10 which is the population mean
#the more reps we do, the closer we will get to the population mean - in theory
#the mean of the sampling distribution for the mean should be a strong estimate for the true mean of the population
#the standard deviation of this distribution will represent the standard error of the sample??? 

#we could also do this with other statistics for the sample (median, mode, etc.)
samp_dist_median <- do(reps) * median(rnorm(n=1000, mean = 10, sd = 2))
str(samp_dist_median)
histogram(samp_dist_median$median)
#^ generates a sample distribution of medians

#one more for standard deviation
samp_dist_sd <- do(reps) * sd(rnorm(n=1000, mean = 10, sd = 2))
histogram(samp_dist_sd$sd)
```

## Standard Error

So, how reliable are these estimates of a population parameter based on the mean of the sampling distribution for a statistic? How far off is a statistic we calculate from the true statistic?

standard error (SE) = square root of the variance of the sampling distribution = standard deviation of the sampling distribution

$$\frac{s}{\sqrt{n}}$$

```{r}
# se = sd(x)/sqrt(length(x))
se_mean <- sd(sdm$mean)
se_median <- sd(samp_dist_median$median)
```

Estimating Standard Error in 3 Ways

1.  standard dev of pop / sqrt (sample size) = expected standard error of a sample from that population

2.  g

3.  g

# Class 11: Distributions

## Playing with Distributions

Draw 100 random numbers from a normal distribution with mean = 2 and sd = 4 - What is the mean of those numbers? - What is the sd? - What is the se of the mean based on this sample? - What is the expected standard error for a sample of this size? - If we draw samples 1000 times of this size and take the mean to get a sampling distribution for the mean, what is the standard error calculated as the standard deviation of this distibution? - What if sample size is changed to 1000? - What if we change it to 20?

```{r}
s <- rnorm(n=100, mean = 2, sd = 4)

mean(s) #2.759087
sd(s) #4.110515
sd(s)/sqrt(length(s)) #0.4110515 -> standard error of this sample
4/sqrt(length(s)) #0.4 -> expected standard error (4 is the sd we have defined)

reps <- 1000
sd_mean <- do(reps) * mean(rnorm(n=100, mean = 2, sd = 4))
#sd_mean$mean will be a vector 1000 elements long where each cell is a mean of one sample
sd(sd_mean$mean) #0.4132877 <- this is standard error?? 

#change sample size to 1000
sd_mean_1000 <- do(reps) * mean(rnorm(n=1000, mean = 2, sd = 4))
sd(sd_mean_1000$mean) #0.1274715

sd_mean_20 <- do(reps) * mean(rnorm(n=20, mean = 2, sd = 4))
sd(sd_mean_20$mean) #0.8766896
```

Now, let's try it with a t distribution. Lower numbers of df (lower sample size) will output a shorter, flatter dist. Lower kurtosis.

Step 1: draw a t distribution

```{r}
#ignoring non-centrality parameter (default = 0)
#changing degree of freedom parameter
plotDist("t", df = 100, xlab = "x", ylab = "Frequency", col = "red")
plotDist("t", df = 50, add = TRUE)
plotDist("t", df = 25, add = TRUE)
plotDist("t", df = 12, add = TRUE)
plotDist("t", df = 6, add = TRUE)
plotDist("t", df = 3, add = TRUE)
plotDist("t", df = 1, add = TRUE)
```

Above df = 30, t dist and normal dist are basically the same shape.

Step 2: Draw 100 random numbers from a t-dist with 99 degrees of freedom and a non-centrality at 2

```{r}
st <- rt(n=100, ncp = 2, df = 99)

mean(st) #2.056604
sd(st) #0.9839444
sd(st)/sqrt(length(st)) #0.09839444 -> standard error of this sample

sd_mean_st <- do(reps) * mean(rt(n = 100, df = 99, ncp = 2))
glimpse(sd_mean_st)
histogram(sd_mean_st$mean)
```

Now, with a beta distribution. Parameters are a (shape 1) and b (shape 2)

```{r}
plotDist("beta", shape1 = 2, shape2 = 4, xlab = "x", xlim = c(0,1), ylab = "Frequency", col = "red")
plotDist("beta", shape1 = 1, shape2 = 1, add = TRUE) #flat line
plotDist("beta", shape1 = 0.5, shape2 = 0.5, add = TRUE) #u shape
plotDist("beta", shape1 = 2, shape2 = 0.5, add = TRUE) #exp growth looking
plotDist("beta", shape1 = 0.5, shape2 = 2, add = TRUE) #exp dec looking
plotDist("beta", shape1 = 4, shape2 = 2, add = TRUE) #backwards of original
```

Using shape1 = 0.3, shape2 = 4

```{r}
sb <- rbeta(n=100, shape1 = 0.3, shape2 = 4)

mean(sb) #0.070244
sd(sb) #0.09899034
sd(sb)/sqrt(length(sb)) #0.009899034 -> standard error of this sample

sd_mean_sb <- do(reps) * mean(rbeta(n = 100, shape1 = 0.3, shape2 = 4))
sd(sd_mean_sb$mean) #0.01053539 = estimated standard error for the pop
histogram(~mean, data = sd_mean_sb) #looks normal along the mean
#the sampling distribution is always normally distributed, regardless of the distribution you pull samples from 
```

## Key Functions for Distributions

r\_(n = ) draws random samples of size n from a given standard distribution

p\_(q = ) returns the quantile associated with a given value of q

q\_(p = ) returns the value of p at a given quantile through the distribution

d\_(x = ) returns value of the probability density function (pdf) at the value of x

```{r}
#How far through the distribution to we have to go to get to x = q? 
pnorm(q = 0, mean = 0, sd = 1)
#0.5, where q is 0 i am halfway (50%) along the normal dist

#what value of x is associated with being halfway along a distribution? 
qnorm(p = 0.5, mean = 0, sd = 1)
#0, where q is 0 p = 0.5
```

What values of x are associated with being 0.025 and 0.975 quantiles (aka the boundaries of 95% of the distribution)?

```{r}
qnorm(p = c(0.025, 0.975), mean = 0, sd = 1) #need a known mean and sd
#[1] -1.959964  1.959964
#this expresses the boundaries of the central 95% of a given distribution
#a CI is -1.96 to 1.96 standard deviations from the mean

#what if we hvae a beta dist? 
qbeta(p = c(0.025, 0.975), shape1 = 2, shape2 = 4)
#[1] 0.05274495 0.71641794
#this dist is skewed, so 95% of data is within 0.5 and 0.71
```

```{r}
lower <- qbeta(0.025, shape1 = 2, shape2 = 4)
upper <- qbeta(0.975, shape1 = 2, shape2 = 4)
curve(dbeta(x, shape1 = 2, shape2 = 4), from = 0, to = 1, lwd = 2) #drawing a beta distribution with given shapes, fit between 0 and 1
x <- seq(lower, upper, length = 500)
polygon(c(lower, x, upper),
        c(0, dbeta(x, shape1 = 2, shape2 = 4), 0),
        col = "lightblue")
abline(v = c(lower, upper), col = "red", lwd = 2)

#how much of the information is to the left of the lower bound? 
pbeta(q = lower, shape1 = 2, shape2 = 4) #0.025
#yay it works! replicable back and forth :) 

#how much of the information is to the left of 0.2? (as an example)
pbeta(q = 0.2, shape1 = 2, shape2 = 4) #0.26272, the point x=0.2 is 26% of the way through the dist

#what is the expected median for a sample drawn from this dist? 
qbeta(p = 0.5, shape1 = 2, shape2 = 4) #0.3138102, at the point 0.313, 50% of the data lies on either side
```

## Confidence Intervals

An interval around our estimated mean for the sampling distribution, giving us a range around that mean that any given sample from the population would be some amount of likely to fall within.

A 95% CI, for instance, tells us that 95% of the time, a sample taken from a sample distribution, would have a mean within the CI range.

CI = point estimate +\_ critical value X SE

$$ CI = estimate \pm {critical value * SE} $$

Using a normal distribution:

```{r}
ci.n <- mean(x) + qnorm(c(0.025, 0.0975)) * (sd(x)/sqrt(length(x)))
```

Example - calculating a theoretical CI:

```{r}
x <- c(2.9, 4.8, 8.9, -3.2, 9.1, -2.5, -0.9, -0.1, 2.8, -1.7)
m <- mean(x)
se <- sd(x) / sqrt(length(x))

ci <- m + qnorm(c(0.025, 0.975)) * se
ci
#we are assuming that the mean of any number of values pulled from vector x will be normally distributed -> the mean of any 10 numbers pulled from x should fall within the CI 95% of the time
```

## Central Limit Theorem

The sampling distribution of averages (or any other summary statistic) of "independent and identically distributed" variables approaches a normal distribution as sample size increases.

This allows us to estimate summary statistics at a population level even though we can only observe a sample of the true data that exists.

Regardless of the underlying probability distribution of a population of iid random variables, the distribution of a statistic should:

-   be approximately normal

-   be centered at the population mean for the statistics

-   have ##CHECK NOTES##

## Small Sample Sizes - t Distributions

When sample size is \< 30 or so, we don't use the normal dist. to calculate CIs. Can use Gosset's or Student's T distribution.

A t distribution is a continuous probability distribution that is similar in shape to normality, but has fatter tails at smaller sample sizes. Sample size is input as degrees of freedom (df = sample size - 1) to create the distribution for a given sample.

Using a t distribution:

```{r}
ci.t <- mean(x) + qt(c(0.25, 0.975), sf = length(x) - 1) * (sd(x)/sqrt(length(x)))
```

## Bootstrapping

We do not imagine that the sampling distribution has a particular shape (normal, t).

Instead, we create a distribution by resampling from our own sample many many times using a simulation process known as Monte Carlo. This allows us to approximate a sampling distribution without needing the entire population data or making assumptions about the distribution of critical points.

To generate this we need to \[1\] resample from our observed data with replacement many times (same number of observations as original sample), \[2\] calculate our statistic of interest for each resample, \[3\] plot all of those statistics of interest to create a sampling distribution that is specific to our data.

We can use the quantile() function (like qnorm, qt, etc.) to define the limits of the CI based on the central tendency of our many sample statistics.

```{r}
#set values
n_boot <- 10000 #how many times to resample
boot <- vector()
n <- length(x)

#using a for loop to create boot
for(i in 1:n_boot){
  boot[[i]] <- mean(sample(x, n, replace = TRUE))
}

#alternative to for loop using do()
#boot <- do(n_boot) * mean(sample(x, n, replace = TRUE))

#final step v
ci.b <- quantile(boot, probs = c(0.025, 0.975))
ci.b
```

# Class 12: Hypothesis Testing

H0, null hypothesis Ha, alternative hypothesis(-es)

To implement a hypothesis test, we need to:

-   calculate a test statistic based on our data

-   calculate the p value associated with the test stat (probability of obtaining, by chance, a test stat that is as hgih or higher than the calculated one)

-   compare test stat to an appropriate sampling distribution to yield a p value

-   compare p value with significance level ($\alpha$)

In parametric statistics, we make the assumption that the sampling distribution will mimic a known distribution (normal, t, etc.).

$$ Z = \frac {\bar x - mu} {s / \sqrt{n}} $$ \# Example: One Tailed T Test Adult woolly monkeys have average body weight of 7.2kg. In another population, which may be a different species, I weigh 15 animals. I want to know if the weight of the second population is meaningfully different than in the first group.

```{r}
library(tidyverse)
x <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/woolly-weights.csv", col_names = TRUE)

colnames(x)
mean(x$weight) #mean is 6.4. Could this mean have been pulled from a the first population? 

m <- mean(x$weight)
sd <- sd(x$weight) #0.89
se <- sd/sqrt(length(x$weight)) #0.23

#our expectation (null hyp) is that mean - mu = 0 (no difference)

mu <- 7.2 #expectation
t_stat <- (m - mu) / se
t_stat #-3.3

# What is the 95% CI around our mean estimate? 
# If it doesn't include 7.2, that would be an early indicator that t might be significant.
boot <- vector()
for(i in 1:1000){
  boot[[i]] <- mean(sample(x$weight, length(x$weight), replace = TRUE))
}
ci.b <- quantile(boot, probs = c(0.025, 0.975))
ci.b #5.977, 6.87
# this CI does NOT include the expected mean
# this means we have pretty good evidence that the sample is drawn from a different population than our expectation came from
```

## Two-Tailed T Test

What is the two-tailed probability of getting a statistic as far away from the expected value (0) as it is BY CHANCE?

We need to calculate the chance it is higher + the chance it is lower (two tails).

```{r}
#t_stat is test statistic (= - 3.3), mean of sample is 6.3
#what is the chance of getting a test statistic that is >-3.3 or <3.3
p_lower <- pt(-1 * abs(t_stat), df = n - 1) #gives us the area under the curve (t distribution) for values <-3.3
p_upper <- 1 - pt(1 * abs(t_stat), df = n - 1)
p <- p_lower + p_upper
p #0.008704304

#what is the 95% CI around the mean for this sample? 
ci <- m + qt(p =(0.025, 0.975), ncp = 0, df = n-1) * se

#using the t test function in R
t.test(x$weight, mu = 7.2)
```

T statistic for a two sampled t test assuming unequal variance between groups

$$ T = \frac{\bar x2 - \bar x1 - \mu}{\sqrt(\frac{s1^2}{n1} + \frac{s2^2}{n2})} $$ Where $\mu$ is expected difference between samples according to null hyp (usually 0), s is variance for each sample, n is number of observations for each sample, and xbar is the actual mean of each sample

Degrees of freedom for a two sampled t test assuming unequal variance between groups

***Check Class Challenge for an example of a two sample T test***

## Permutation-Based Inference

1.  What should we use as a test statistic?

2.  What is the null hypothesis?

3.  What is the alternative hypothesis?

4.  Generate a permutation distribution for the test statistic of interest.

5.  Determine the P Value.

To demonstrate, we will use the data set from class challenge "Two Sample T Test." We are interested in the difference in home range size between males and females.

```{r}
#setup
d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/tbs-2006-2008-ranges.csv")
head(d)
d <- d |> select(id, sex, sex.code, kernel95)
```

```{r}
#What is the actual difference between the means? 
obs_diff <- d |> 
  summarize(diff = mean(kernel95[sex == "M"]) - mean(kernel95[sex == "F"])) |>
  pull(diff)
obs_diff #110.9561
```

```{r}
#What is the permuted difference between means? 
#create a permutation distribution and see if our obs_diff falls within it and/or if our null hyp falls within it
n_perm <- 10000
perm_diff <- vector() #will hold the results of each permuted mean in a vector
for (i in 1:n_perm){
  d |> 
    mutate(sex = sample(sex)) |> #shuffle sex randomly within the data set, default replace = FALSE to make sure the number of M and F are the same as original
    summarize(diff = mean(kernel95[sex == "M"]) - mean(kernel95[sex == "F"])) |>
    pull(diff) -> perm_diff[[i]] #pull out result to put it into the dummy vector
}
histogram(perm_diff) #appears normal
#interestingly, the observed difference is at the very top end of the permuted distribution
#if our null hyp is correct, the difference would be 0
#0 is close to the midpoint of this histogram, which suggests we could accept the null, but we need to do some more testing to confirm
```

Permutation testing like this makes no assumptions about the distribution of the original data or the sampled data. This is a robust system that is not dependent on sample size or having known distribution patterns. NO DOWNSIDE... but it takes longer than parametric tests for computers to do. Modern computers will have no problem doing it at pretty large scales.

```{r}
#Alternative methods to create a permutation distribution
### ADD LATER

#do()
perm_diff <- {do(n_perm) * d |> 
    mutate(sex=sample(sex))}

#replicate()
```

```{r}
#Computing a P value
#How many times is our permuted value bigger than our observed value? 
p_val <- sum(abs(perm_diff) >= abs(obs_diff))/length(perm_diff)
p_val #9e-04

#alternative 
p_val <- mean(abs(perm_diff) >= abs(obs_diff))
```

The {infer} package offers functions to do permutation tests without needing to code them yourself! Works whether we are dealing with means, difference of means, proportions, difference of proportions.

```{r}
install.packages("infer")
library(infer)
```

```{r}
#use specify() to indicate the variables we are interested in
d <- d |> 
  specify(formula = kernel95 ~ sex)

#use hypothesize() to declare H0
d <- d|> hypothesize(null = "independence")
#"independence" means there is no relationship, variables operate independently of one another
#null could be different for a one-tail test, for instance

#use generate() to generate replications of permuted data under the assumption that null is true
perm <- d |> generate(reps = 1000, type = "permute")

#use calculate() to present summary statistics of interest for each replication
perm_diff <- perm |> calculate(stat = "diff in means", order = c("M", "F"))
#perm_diff is a df that tracks perm number and diff of means (as "stat")

#use visualize() to show the null distribution (the permuted distribution)
visualize(perm_diff, bins = 20)
#NOT WORKING

#use visualize() to show the test statistic relative to the permuted distribution
obs_diff <- d |> specify(kernel95 ~ sex) |> calculate(stat = "diff in means", order = c("M", "F")) #just gives the obs diff using infer code

visualize(perm_diff, bins = 20) +
  shade_p_value(obs_stat = obs_diff, direction = "both")

get_p_value(perm_diff, obs_diff, direction = "both")
#returns a warning about p = 0 because NONE of the permutations had a diff of means greater than the observed difference
```

# Class 13: Intro to Regression

```{r}
install.packages("lmodel2")
install.packages("sjPlot")
install.packages("broom")
install.packages("manipulate")
install.packages("patchwork")

library(tidyverse)
library(manipulate)
library(patchwork)
library(infer)
```

Basic premise of modeling is to explore a relationship between... - an outcome/response/dependent variable, **y** - one or more epxlanatory/predictor/independent variables, **x**

**Simple Linear Regression** (general)

-   outcome is a continuous numerical variable

-   single predictor that is either categorical or numerical

**Multiple Linear Regression** (general)

-   outcome is a continuous numerical variable

-   more than one predictor variable, either categorical or numerical

**ANOVA/ANCOVA**

-   focus on categorical predictors

-   ANOVA = one or more categorical independent variables

-   ANCOVA = a combination of categorical and continuous independent variables

**Generalized Linear Regression**

-   allows for binary, categorical, count variables as outcomes

**Descriptive Statistics**

Covariance: between 2 variables is the product of the deviation of each variable from their respective means divided by df (n-1)

cov(x,y) = SUM ((x - xbar)(y - ybar)) / (n-1)

Correlation Coefficient: standardized covariance - divided by product of standard deviations

cor(x, y) = cov(x,y) / sd(x)sd(y)

```{r}
d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/zombies.csv", col_names = TRUE)
names(d)

#What is the covariation between weight and height? 
cov_wh <- sum(((d$weight - mean(d$weight))*(d$height - mean(d$height)))/(nrow(d) - 1))
cov_wh #66.03314
cov(d$height, d$weight) #same! it worked! 

#What is the correlation coefficient between weight and height? 
cor_wh <- (sum(((d$weight - mean(d$weight))*(d$height - mean(d$height)))/(nrow(d) - 1))) / (sd(d$weight)*sd(d$height))
cor_wh #0.8325862
cor(d$weight, d$height) #same! works! 

#plot the relationship between variables
plot(d$weight ~ d$height)
```

Regression can be used to...

-   use one of more variables to predict the value of an outcome based on information contained in predictor variables
-   explicity describe and quantify the relationship between x and y, determine the significance of this relationship, generate measures to summarize the relationship, and identify causal relationships (sometimes)
-   develop and choose among different models of the relationship between two variables
-   do analyses of covariation among sets of variables to identify their relative explanatory power (what x is best able to predict y)

Regression formula:

Typically, we model y as a linear function of the variables. yi = beta0 + beta1\*Xi + epsiloni

beta0 = intercept B1 = slope ep = each obervation is a deviation away from the line (the error term)

The beta values are "regression coefficients" that our analysis is trying to estimate, while minimizing the error term. This leaves us basically with $y=mx+b$, allowing us to predict y for a given x.

**Ordinary Least Squares**: minimizes the residuals (AKA the squared difference of each y.obs from the predictive line)

```{r}
temp <- d |> mutate(centered_height = height - mean(height),
                    centered_weight = weight - mean(weight))

slope.test <- function(beta1, data) {
  g <- ggplot(data = data, aes(x =centered_weight, y = centered_height))
  g <- g + geom_point()
  g <- g + geom_abline(intercept = 0, slope = beta1, alpha = 1/2)
  ols <- sum((data$centered_height - beta1 * data$centered_weight)^2)
  g <- g + ggtitle(paste("Slope +", beta1, "\nSum of Squared Deviations =", round(ols, 3)))
}

manipulate(slope.test(beta1, data = temp), beta1 = slider(-1, 1, initial = 0, step = 0.005))
#this would in theory show how least squares regression works with a slope you can move on your own
```

```{r}
head(d)
m <- lm(height ~ weight, data = d) #creating a linear model of this relationship
summary(m)
#give estimate of intercept, slope
```

Exercise: 
- Plot testis size in relation to male body size, what is that relationship like? 
- Calculate beta0 and beta1 for the regression of: log(combined_testis_mess) ~ log(body_mass_male_mean)
- run the lm function and summarize to compare 

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/comparative_primate_sexuality_data.csv"
d <- read_csv(f, col_names = TRUE)
names(d)

ggplot(data = d, aes(x = log(Body_mass_male_mean), y = log(Combined_testis_mass))) +
  geom_point(na.rm = TRUE)

temp <- d |> filter(!is.na(Combined_testis_mass))

beta1 <- cov(log(temp$Body_mass_male_mean), log(temp$Combined_testis_mass))/ var(log(temp$Body_mass_male_mean))
beta1 #0.75

##GO BACK AND ADD
m <- lm(formula = log(Combined_testis_mass) ~ log(Body_mass_male_mean), data = temp)
summary(m) #yay same number :) 

names(m) 
#m$coefficients holds regression coefficients
#m$model holds original x and y values
#m$fitted.values holds fitted values for y = yhat
#m$residuals holds residuals for each obesrvation = y - yhat
#broom :: tidy(m) gives a clean table of beta coeff, p values, intercepts etc.
#broom :: glance(m) gives summary statistics for the model as a whole
broom::tidy(m)
#statistic = estimate / standard error
```

Alternatives to OLS regression: 

```{r}
library(lmodel2)

m <- lm(formula = log(Combined_testis_mass) ~ log(Body_mass_male_mean), data = temp)
summary(m)

m2 <- lmodel2(log(Combined_testis_mass) ~ log(Body_mass_male_mean), data = d, 
              range.y = "relative", range.x = "relative", nperm = 1000)
summary(m2)
##NOT WORKINGGGGG

```